A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-02-22 paper/writeup.tex
figure-6changes (hopefully final)

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}

@@

Concretely, our method can be applied to any two tree sequences,
as illustrated in \figref{fig:conceptual_discrepancy}.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(n_1, n_2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
Further, we define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(n_1, n_2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
The \emph{matched span} of $\T_1$ in $\T_2$ is then defined to be
\begin{align*}
    \match(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, 
and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
its \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align}
    \match(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align}
If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\match(\T_1, \T_2)$.
Let $s(\T,n)$ denote the total span that node $n$ is present in the local trees,
\begin{align*}
	s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise),
and let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{non-matched span} of $\T_1$ in $\T_2$ by
\begin{align*}
	\nmatch(\T_1,\T_2)
	=
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\match(\T_1,\T_2),
\end{align*}
which is the total span for all nodes in $\T_1$
over which their descendant samples do \emph{not} match
 those of their best match in $\T_2$.
Contrarily, given a matching $\beta:\T_1\to\T_2$, 
we want to quantify how much of $\T_2$ is represented in $\T_1$.
To do this, we define the \emph{inverse matched span} of $\T_1$ in $\T_2$ as
\begin{align}
	\invmatch(\T_1,\T_2) = \sum_{n_2\in N_2} \max_{n_1\in \alpha^{-1}(n_2)} m(n_1,n_2)
\end{align}
where $\alpha^{-1}(n_2)$ is the set of all nodes $n_1\in \T_1$ 
whose best match is $n_2$.
Note that the inverse matched span is not equivalent
 to the matched span of $\T_2$ in $\T_1$ 
 as the inverse matched span maximizes over spans
  which are still dependent on $\beta$, 
  and, as previously stated, $\beta$ is not guaranteed to be bijective.

@@

\begin{figure}[!hbt]
	\begin{center}
        \includegraphics[width=\textwidth]{figures/figure6-ARF-TPR-plt.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (right)
        and sequence lengths, with 1,000 diploid samples (left).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified} (`S'), removing the unary spans,
        and (ii) \emph{inferred} (`I'), using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended} (`SE', `IE').
        The inferred, then simplified, ARG ('IS') and its subsequent extension (`ISE') are also shown.
        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (\eqnref{eqn:dissimilarity}).
        \textbf{Bottom row:} TPR, agreement to the true ARG
        as proportion of the true ARG that is represented.
        }     
    \label{fig:dissimilarity}
\end{figure}

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
As illustrated by~\cite{wong2024general}, there is a significant diversity
in the structures inferred by current methods,
and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
infers ARGs containing unary nodes as a byproduct of its inference algorithm.
% Note this seems pretty weak and is sure to be dinged. We could at least 
% comment on what we might observe with other inference methods? Would we 
% recommend that people extend_edges() for their SINGER or Relate ARGs?
A comparison across other ARG inference methods is left for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 2\% (bottom panels),
but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span ($\approx <1\%$).
Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.)

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\tsdate}{\texttt{tsdate}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}

@@

Concretely, our method can be applied to any two tree sequences,
as illustrated in \figref{fig:conceptual_discrepancy}.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(n_1, n_2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
Further, we define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(n_1, n_2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
The \emph{matched span} of $\T_1$ in $\T_2$ is then defined to be
\begin{align*}
    \match(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, 
and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
its \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align}
    \match(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align}
If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\match(\T_1, \T_2)$.
Let $s(\T,n)$ denote the total span that node $n$ is present in the local trees,
\begin{align*}
	s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise),
and let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{non-matched span} of $\T_1$ in $\T_2$ by
\begin{align*}
	\nmatch(\T_1,\T_2)
	=
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\match(\T_1,\T_2),
\end{align*}
which is the total span for all nodes in $\T_1$
over which their descendant samples do \emph{not} match
 those of their best match in $\T_2$.
Contrarily, given a matching $\alpha:\T_1\to\T_2$, 
we want to quantify how much of $\T_2$ is represented in $\T_1$.
To do this, we define the \emph{inverse matched span} of $\T_1$ in $\T_2$ as
\begin{align}
	\invmatch(\T_1,\T_2) = \sum_{n_2\in N_2} \max_{n_1\in \alpha^{-1}(n_2)} m(n_1,n_2)
\end{align}
where $\alpha^{-1}(n_2)$ is the set of all nodes $n_1\in \T_1$ 
whose best match is $n_2$.
Note that the inverse matched span is not equivalent
 to the matched span of $\T_2$ in $\T_1$ 
 as the inverse matched span maximizes over spans
  which are still dependent on $\beta$, 
  and, as previously stated, $\beta$ is not guaranteed to be bijective.

@@

\begin{figure}[!hbt]
	\begin{center}
        \includegraphics[width=\textwidth]{figures/figure6-ARF-TPR-plt3.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (right)
        and sequence lengths, with 1,000 diploid samples (left).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified} (`S'), removing the unary spans,
        and (ii) \emph{inferred} (`I'), using \tsinfer{} on genotypes
        and dated using \tsdate{};
        then each of these had its haplotypes \emph{extended} (`SE', `IE').
        The inferred, then simplified, ARG ('IS') and its subsequent extension (`ISE') are also shown.
        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (\eqnref{eqn:dissimilarity}).
        \textbf{Bottom row:} TPR, agreement to the true ARG
        as proportion of the true ARG that is represented.
        }     
    \label{fig:dissimilarity}
\end{figure}

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
As illustrated by~\cite{wong2024general}, there is a significant diversity
in the structures inferred by current methods,
and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
infers ARGs containing unary nodes as a byproduct of its inference algorithm.
% Note this seems pretty weak and is sure to be dinged. We could at least 
% comment on what we might observe with other inference methods? Would we 
% recommend that people extend_edges() for their SINGER or Relate ARGs?
A comparison across other ARG inference methods is left for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
Since our matched spans method breaks ties with time, 
we date the re-inferred ARG using \tsdate.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 4\% (bottom panels),
but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span ($\approx <1\%$).
Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.) 
We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Figure-6}.

@@

\begin{table}[!h]
	\begin{center}
		\begin{tabular}{{|c||c|c|c|c|c||c|c|c|c|c|}}
			\hline
			& \multicolumn{5}{|c||}{ARF} & \multicolumn{5}{|c|}{TPR}\\
			\hline
			\hline
			Sample Size   & 10 & 50 & 100 & 500 & 1000 & 10 & 50 & 100 & 500 & 1000\\
			\hline\hline
			I & 0.3808 & .3490 & 0.3218 & 0.2734 & 0.2552 & 0.2904 & 0.2873 & 0.2908 & 0.3099 & 0.3299\\
			\hline
			IE & 0.4518 & 0.3932 & 0.3584 & 0.2932 & 0.2726 & \textbf{0.2917} & \textbf{0.2885} & \textbf{0.2916} & \textbf{0.3110} & \textbf{0.3309}\\
			\hline
			IS & \textbf{0.1993} & \textbf{0.1802} & \textbf{0.1701} & \textbf{0.1577} & \textbf{0.1516} & 0.2717 & 0.2703 & 0.2732 & 0.2909 & 0.3109\\
			\hline
			ISE & 0.3129 & 0.2779 & 0.2587 & 0.2332 & 0.2189 & 0.2850 & 0.2819 & 0.2848 & 0.3023 & 0.3228\\
			\hline
			\hline
			Length   & $1\textnormal{e}6$ & $5\textnormal{e}6$ & $1\textnormal{e}7$ & $3\textnormal{e}7$ & $5\textnormal{e}7$ & $1\textnormal{e}6$ & $5\textnormal{e}6$ & $1\textnormal{e}7$ & $3\textnormal{e}7$ & $5\textnormal{e}7$\\
			\hline\hline
			I & 0.1383 & 0.2167 & 0.2324 & 0.2525 & 0.2595 & 0.5259 & 0.4478 & 0.4108 & 0.3540 & 0.3317\\
			\hline
			IE & 0.1474 & 0.2289 & 0.2469 & 0.2702 & 0.2754 & \textbf{0.5263} & \textbf{0.4485} & \textbf{0.4117} & \textbf{0.3551} & \textbf{0.3326}\\
			\hline
			IS & \textbf{0.0948} & \textbf{0.1359} & \textbf{0.1407} & \textbf{0.1485} & \textbf{0.1526} & 0.5172 & 0.4331 & 0.3939 & 0.3356 & 0.3118\\
			\hline
			ISE & 0.1193 & 0.1827 & 0.1960 & 0.2140 & 0.2216 & 0.5232 & 0.4432 & 0.4051 & 0.3481 & 0.3234\\
			\hline
		\end{tabular}
		\caption{Values for 'I', 'IE', 'IS', 'ISE' in \figref{fig:dissimilarity}.}
		\label{tab:Fig-6}
	\end{center}
\end{table}
